---
title: "Introduction to RNAseq"
author: Hirad Hosseini
date: 07/24/22
---

```{r setup}
    knitr::opts_knit$set(root.dir = normalizePath("../week5_more_clinical/analysis_data/")) 
```

```{r}
# install, only if needed, and load packages (SummarizedExperiment and TCGAbiolinks)
install.packages("maftools")
install.packages("SummarizedExperiment")
```

Make sure you have access to rna_se. If it is not already in your workspace (ie already queried, downloaded, and prepared), let Nicole know NOW so we can get you the data a different way.

###### (1) Summarized Experiment Basics

Let's take a look at rna_se (RNA Summarized Experiment). Call head() and str(). Also explore the data frame by double clicking on it from the Environment panel (or calling "View(rna_se)").

```{r Exercise 1.1}
# write code here
head(rna_se)
str(rna_se)
```

Use the dim() function to find the dimensions of the colData, rowRanges, and unstranded dataframes. Record them below:

  ~ colData (rows, cols): 1226, 87
  ~ rowRanges: 60660, 10
  ~ unstranded: 60660, 1226

```{r Exercise 1.2}
# write code here
dim(rna_se@colData)
dim(rna_se@rowRanges@elementMetadata)
dim(rna_se@assays@data$unstranded)
```

Run the following code and fill out a description of the rows/columns of our data frames.You can use this as a cheat sheet moving forward.
  ~ colData
        - Rows: patient barcodes
        - Columns: info about the patient
  ~ rowRanges
        - Rows: genes themselves
        - Columns: info about the genes
  ~ unstranded (in assays)
        - Rows: genes
        - Columns: patients
  
```{r Exercise 1.3}
# colData: 
rownames(rna_se@colData)
colnames(rna_se@colData)

# rowRanges:
str(rna_se@rowRanges)
# what are the rows... use your logic

# for unstranded, don't call anything, just use your logic from exercise 1.2 and and think it through!
```

###### (2) ColData

View the first 15 values in the age_at_index column.

```{r Exercise 2.1}
# write code here
rna_se@colData$age_at_index[1:15]
```

Call summary on the age_at_index column. Record the output below:
    Min.: 26
    1st Qu.: 48
    Median: 58
    Mean: 58.27
    3rd Qu.: 67
    Max.: 90
    NA's: 1
    
```{r Exercise 2.2}
# write code here
summary(rna_se@colData$age_at_index)
```

Create a data frame called rna_clinical that contains data from colData, except for the NA patient and their data.

```{r Exercise 2.3}
rna_clinical <- rna_se@colData$age_at_index[!is.na(rna_se@colData$age_at_index)]
```

Use the as.data.frame() function to change this from an S4 object to a data frame object.

```{r Exercise 2.4}
rna_clinical <- as.data.frame(rna_clinical)
```

Make a mask and subset out the treatments column.

```{r Exercise 2.5}
treatments_mask <- ifelse(colnames(rna_se@colData) == "treatments" | colnames(rna_se@colData) == "disease_type" | colnames(rna_se@colData) == "primary_site", FALSE, TRUE)
sum(treatments_mask)
rna_clinical <- rna_se@colData[treatments_mask]
```

Save rna_clinical to your local machine

```{r Exercise 2.6}
write.csv(rna_clinical, "../../week7_rna/brca_rna_clincial_data.csv", row.names = TRUE)
```

###### (3) RowRanges

Call head() on rowRanges@elementMetadata:

```{r Exercise 3.1}
# write code herej
head(rna_se@rowRanges@elementMetadata)
```

Create a data frame called rna_genes that contains data from elementMetadata.

```{r Exercise 3.2}
rna_genes <- rna_se@rowRanges@elementMetadata
```

Use the as.data.frame() function to change this from an S4 object to a data frame object

```{r Exercise 3.3}
rna_genes <- as.data.frame(rna_genes)
```

Save rna_genes to your local machine.

```{r Exercise 3.4}
write.csv(rna_genes, "../../week7_rna/brca_rna_gene_data.csv", row.names = FALSE)
```

###### (4) Assays

What are the names of the 6 assays? unstranded stranded_first stranded_second tpm_unstrand fpkm_unstrand fpkm_uq_unstrand


```{r Exercise 4.1}
# access the assays here
rna_se@assays
```

Create a new data frame called "rna_counts" that contains the data from unstranded. Recall the patient with NA age data? We need to exclude that patient. We can use the same mask to do this.

```{r Exercise 4.2}
rna_mask = ifelse(is.na(rna_se@colData$age_at_index) == TRUE, FALSE, TRUE)
rna_counts <- rna_se@assays@data$unstranded[,rna_mask]

```

Fetch the data for the first 5 patients and the first 15 rows.

```{r Exercise 4.3}
# write code here
#rna_counts[1:5,]
rna_counts[1:15, 1:5]
```

Save rna_counts to your local machine (warning: this will likely take a while!)

```{r Exercise 4.4}
write.csv(rna_counts, "../../week7_rna/brca_rna_count_data.csv", row.names = TRUE)
```
